### 2017-02-07 manuscript.rmd
wor smithing, added tau v max expression plot

With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species. Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future funcitonal genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees. The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendancy of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise). There are several difficulty with interpretting this plot, though. First, it does not just reflect the evolutionary porcesses that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

These simulations have several implcations. The pairwise comparisons used by KMRR cannot distinguish betweeh the null hypothesis and ortholog conjecture, as the expected results are the same under both hypotheses. They also serve as a validation of phylogenetic methods as applied to this problem. Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture. In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hyptheses about evolutionary process.

@@

Once we addressed the issues above with the implementation of pairwise comparisons in Levin *et al.* [@Levin:2016bp], we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species. We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure XXLevin b). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches (Figure 1a). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

@@

Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species. Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics stidues. There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies. One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined. In comparative functional genomics analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits. This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa]. In addition to methods development, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.

@@

This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].

@@

The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods. Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.

These analyses illustrate how important it is to not conflate evolutionary patterns with the processes that generated them. Finding a pattern wherein paralogs tend to be more different than orthologs is not evidence that there are different processes by which orthologs and paralogs evolve. This is also the expected pattern when they evolve under the same process but paralogs tend to be more distantly related to each other than orthologs are. The fact that multiple pairwise comparisons of developmental gene expression across diverse species share a particular pattern is not evidence of a general process that explains the differences between all species in the analysis. It is also the expected pattern when there are unique changes specific to a single species, as these changes will impact all pairwise comparisons to that species and therefore be counted multiple times. To use patterns across living species to test hypotheses about evolutionary processes, it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies. There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomics traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.

@@

@@

We examined the sensitivity of our results to the specification of calibration dates for the speciation nodes. In `r noised_replicate_n` reanalyses, we drew a new date for each calibration from a normal distribution with the mean of the original date and a standard deviation `r sd_fraction` times the original date. If any daugter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology. The minimum Wilcoxon p in these reanalyses was `r min( as.numeric( unlist( p_noised ) ) )`, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events. This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.

>>>>

With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species. Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees. The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree. We present eight of the actual gene trees with Tau values and independent contrasts in Figure XX_EXAMPLE_TREES as an example of how these methods are applied.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise). There are several difficulty with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

These simulations have several implications. The pairwise comparisons used by KMRR cannot distinguish between the null hypothesis and ortholog conjecture even when the evolution of expression is `r dup_adjust` fold higher following duplication than speciation. The pairwise results are the same under both hypotheses. They also serve as a validation of phylogenetic methods as applied to this problem. Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture. In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hypotheses about evolutionary process.

Their Figure 3 differs from their analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that as paralogs diverge in expression, there is a tendency for the one that evolves to have lower expression also evolve to become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure XX_TAU_V_EXP). A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

@@

Once we addressed the issues above with the implementation of pairwise comparisons in Levin *et al.* [@Levin:2016bp], we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between the same species. We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison) and no support for the inverse hourglass (Figure XXLevin b). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1a). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

@@

Our results highlight the importance of explicitly incorporating information about phylogenetic relationships when comparing functional genomic traits across species. Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to comparative functional genomics studies. There are also interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics studies. One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined. In comparative functional genomics analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits. This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa]. In the same manuscript we also considered another issue of relevance here -- the read counts generated by RNA-seq expression studies cannot be directly compared across species. This is because there are various species-specific technical factors that can be mistaken for differences in expression across species. These can be canceled out within species before making comparisons across species.

@@

@@

The fact that the first two comparative functional genomics studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other studies that use these methods, and that future comparative studies will be compromised if they continue to use pairwise methods. Studies of evolutionary functional genomics should not be focused on the tips of the tree using pairwise comparisons. They should explicitly delve into the tree with phylogenetic comparative methods. Here we show a concrete case where the predictions of a null hypotheses and an alternatives hypothesis do not differ when considering pairwise comparisons, but make distinct predictions with phylogenetic comparative methods that can be used to test them.

These analyses illustrate how important it is to not conflate evolutionary patterns with the processes that generated them. Finding a pattern wherein paralogs tend to be more different than orthologs is not evidence that there are different processes by which orthologs and paralogs evolve. This is also the expected pattern when they evolve under the same process but paralogs tend to be more distantly related to each other than orthologs are. The fact that multiple pairwise comparisons of developmental gene expression across diverse species share a particular pattern is not evidence of a general process that explains the differences between all species in the analysis. It is also the expected pattern when there are unique differences specific to a single species, and the evolutionary changes responsible for these differences are sampled multiple times in pairwise comparisons that span the same phylogenetic branches along which they occurred. To use patterns across living species to test hypotheses about evolutionary processes it is also necessary to incorporate information about evolutionary relationships, *i.e.* phylogenies. There have been decades of work on building comparative phylogenetic methods that do exactly that, and they are just as relevant to comparing functional genomics traits across species as they are to comparing morphology or any of the other traits they are already routinely applied to.

@@

#### Relationship between Tau and maximum expression

```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	nodes_tips = nodes_all %>% filter( ! is.na( species ) )
	max_expression = nodes_tips %>% select( contains("Averaged") ) %>% apply( 1, FUN=max, na.rm=TRUE )
	nodes_tips$max_expression = max_expression

	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
		geom_point(alpha=0.05) + 
		geom_smooth(method="lm") +
		xlab("Maximum Expression") +
		theme_classic()

> Supplementary Figure XX_TAU_V_EXP | There is a negaative relationship between Tau and the maximum observed expression for each gene.

@@

We examined the sensitivity of our results to the specification of calibration dates for the speciation nodes. In `r noised_replicate_n` reanalyses, we drew a new date for each calibration from a normal distribution with the mean of the original date and a standard deviation `r sd_fraction` times the original date. If any daughter nodes became older than their parent, we repeated the sampling until the dates were congruent with the topology. The minimum Wilcoxon p in these reanalyses was `r min( as.numeric( unlist( p_noised ) ) )`, *i.e.* none of them reject the null hypothesis that the rate of evolution of Tau is greater following duplication events than speciation events. This is consistent with the analysis that uses the calibration dates as specified, indicating that our results are robust to the selection of calibration times for speciation nodes.
